\chapter{Ch4-ClosingRemarks}

\section{Experimental tests}
 (microfluidics, red green stuff with tunable overlaps, Gore gut stuff)

The extinction times from the last few chapters are long, and not just those that scale exponentially with the carrying capacity. 
Even the relatively fast results of the Moran model, which scale linearly with $K$, will be longer than is experimentally viable for typical biological populations and timescales. 
The fastest reproducing model organism is the \emph{e. coli} bacterium, which reproduces every twenty minutes in ideal conditions. 
A carrying capacity as low as $10^3$ would show fixation in the Moran limit in a week, but the complete extinction of the population (as described in the first chapter) would take longer than life has existed on the Earth. %NTS:::"first" chapter.
%NTS:::comment on Lenski
%NTS:::somewhere talk about how this isn't a problem, that this is a sort of null model, and works theoretically - if anything is off from this, it allows us to ask pointed "in what way" questions. Minimal working model


\section{Applications of the theory}
%(coalescent theory, phylogenic construction, the obvious of applying it to real biological systems, gut microbiome, ...?)
%\section{Extensions of the theory} - combine with applications
%(eg. would eventually recover Hubbell)

The obvious application of the theories investigated in this thesis is to biological systems of few competing species or strains within constant environments. 
This is somewhat artificial, but there are some systems for which it is relevant. 
%see the previous paragraph for experimental tests
%Any microfluidic work will start with one or two species in a constrained environment. 
Microfluidic work, in addition to its testing capabilities, could model systems with flow, like industrial food processing or the digestive tract. 
The \emph{C. elegans} gut microbiome work of Gore and others \cite{Gore, but also zerbrafish?} suggests that the theories in this thesis could be applicable to microbiota more generally, be they in the gut or lungs or other areas \cite{,,,???}. 
My results rely on a well-mixed assumption, so environments like the gut or the ocean surface are good candidates, but situations like plaque growth or trees(desert scrubs?) are not. %NTS:::this and next sentence
Species need not rely on the environment to mix them well; animals that are mobile or trees with far-travelling seeds would also be candidates. 
%NTS:::more citations above

% (plasmids instead of species)
Chapters 2 and 3 briefly touch on the idea of small population sizes. 
For instance, the second chapter's figure \ref{heatmap} suggests that the exponential term is less relevant than the algebraic term when it comes to the fixation time of two competing species as compared to a similarly sized Moran-like system. 
Indeed, exponential dependence is only dominant when system sizes are large. 
This large population size is relevant in many (if not most) biological contexts, from \emph{e. coli} and their typical $10^6 - 10^8$ bacteria/mL density \cite{Lenski} to hare in Canada's arctic which are more spread out but number in the ten thousands \cite{or whatever}. 
But not all biology is overflowing with individuals. 
One example is that of plasmids in a cell \cite{Ingalls}. 
%
Plasmids are small loops of DNA that typically code for a few/handful genes, often one of which confers some antibiotic resistance. \cite{that bio textbook - Wilson?}
Their reproduction can be thought of as asexual, since only one plasmid copy is required as a template to make a new copy. 
Plasmid copy numbers, the average number of copies of a plasmid per cell, tend to be low, ranging from $10^1$ to $10^3$. 
The copy number can be thought of as the carrying capacity of that plasmid in the cell, and is maintained primarily by a negative regulatory circuit, whereby a protein expressed by the plasmid acts to inhibit the replication of new plasmids. 
There is also a stochastic effect when the cell divides and the plasmids are distributed between the two daughters, but given the differing time scales of plasmid (DNA) replication and cell division it is not a necessity that cellular division be the sole mechanism of local extinction; demographic fluctuations like those described in this thesis may also be relevant. 
Seeing as the bacterial chromosome is typically thousands or a million times longer than that of a plasmid, and as a bacterium can only divide when it has copied its chromosome, I expect the relative time scales to be similarly disparate. 
A low copy number plasmid might have a carrying capacity of $K=20$, which would suggest a mean extinction time of 20 million replications, which is a long time for a bacterium, despite the differing time scales of DNA replication and cell division, though for a quickly replicating, low copy number plasmid in a slowly dividing bacterium this could be of relevance. 
%https://biology.stackexchange.com/questions/31625/when-do-plasmids-replicate-relative-to-its-host-cell-cycle
What's more, different types of plasmids can have the same or similar maintenance mechanisms, such that they compete, as mediated by their shared inhibitor proteins. 
Then the relevant comparison would not be chapter 1's extinction of a single species but the competition of chapters 2 and 3. 
Each plasmid type would have its carrying capacity given by its copy number if it were alone in the bacteria. 
The niche in this case is defined by the replication inhibitor, and niche overlap relates how much the inhibitor of each plasmid type stymies the replication of the other; if there is a shared inhibitor, the plasmids are in the Moran limit, and we expect fixation to be rapid, much faster than the cell division time scale. 

Similar to plasmids, the number of mitochondria in a cell is small and tends to be controlled within a cell. 
In [brewer's] yeast there are typically $34\pm 2$ mitochondria per cell \cite{bionumbers}. 
Of interest to researchers is how the integrity of mitochondrial DNA is maintained \cite{Nunn}. 
Sometimes a mitochondrion will have large deletions in its DNA \cite{???}. 
There are conflicting selection forces in this system: the mutant mitochondria [presumably] reproduce faster than the wildtype but hinder the viability of the cell, hence its reproduction. 
But when one mutant arises in a population, what is the chance it will fixate, and will fixation occur before the cell reproduces? 
The analyses of chapter 3 are relevant to such a problem. 

Coalescent theory is a model that predicts the time in the past when two variants of a gene most recently shared a common ancestor \cite{???}. 
In its simplest form it treats all mutants (at least, all that survived to the present day) as equally viable, interacting with other variants as strongly as they do with themselves \cite{?}. 
In this way it is like a Moran model, or a symmetric Lotka-Volterra model with complete niche overlap. 
%The time it predicts since the last common ancestor is almost exponential in the genetic distance between mutants \cite{}. %incorrect
The time it predicts since the last common ancestor is approximately exponentially distributed, with a mean proportional to the system size \cite{?}. 
An inclusion of incomplete niche overlap, such as has been investigated in chapter 2, would only act to increase the estimated times. 
Niche overlap less than one would be most appropriate for very disimilar variants that now serve different but equally vital purposes in the organism. 
The longer timescale of incomplete niche overlap of course would not change the historical record. 
Rather, coalescent theory is used to make inferences about historical population genetic parameters, like mutation rate and population size. 
For those genes to which incomplete niche overlap applies, the longer timescales are counteracted by a smaller population size or greater mutation rate. 
Conversely, if we know the historic population size and mutation rate, we could infer the niche overlap between disparate gene variants. 
Phylogenetic reconstruction is related to coalescent theory, and would be similarly affected by incomplete niche overlap \cite{???}. 
%NTS:::coalescent theory also discusses the variance of time to most recent common ancestor

%NTS:::FOR HERE AND IN PREVIOUS CHAPTER, READ DYNAMICS SECTION OF WIKIPEDIA UNTB (HUBBELL)
As discussed in the previous chapter, the Hubbell model also relies on assumptions similar to those of the Moran model, to the point that it is effectively a Moran model with immigration \cite{Hubbell2001,others?}. 
It is used to predict the number of species that on average can be found in a system, and the distribution of abundances of the species therein. 
As with coalescent theory, the underlying assumption of Hubbell's neutral theory is that species occupy the same niche. 
This is effectively what he means by the term ``neutral'', a claim which has been controversial \cite{???}. 
Apologists defend the theory by explaining that, in an abstracted sense, species in the same trophic level are effectively competing with each other as much as they compete with themselves \cite{???}. 
The alternative to neutral theory is traditionally taken to be niche theory, the idea that each species in a system occupies its own niche, and any newcomers must either die out as a transient species or invade a niche, suppressing and evicting its previous occupant. 
I have shown, in chapter 3, that there are other alternatives. 
My research investigated the situation where two species have overlapping, but not necessarily identical, niches, such that they do not exclude each other entirely. 
Neither is selected for, and I have focussed on symmetric interactions where each species affects the other to the same degree, has the same carrying capacity, and turnover rate. 
Thus I situate myself among those theories which accommodate both niche and neutral theories \cite{???}. 

%NTS:::ANYTHING ELSE???

%NTS:::paragraph on assumptions/limitations - here, earlier? maybe at end of section, maybe after first paragraph, maybe at start
Of course, some of the assumptions underlying my theories might fail, and this is especially true for systems will small population sizes. 
The results outlined in this thesis should be applied with caution. 
%NTS:::c.f. part in experimental section above where I explain that this is a null model to compare to other theories or from which to base other theories
%well mixed, not spacial
%constant environment
%phenomenology like $r$, $K$, and $a$
%no life stage structure
%lumping all non-focal species together
%where mutant strains come from - clearest for SNPs in genes, but the theory is least applicable to them (unless Moran or selective) - to be symmetric requires two mutations [not true! could increase metabolism of one reactant while decreasing enzymatic activity of another] - SNP can be synonymous or not
%asexual


\section{Conclusions and retrospective}
 (techniques, how to think of niche, competitive exclusion)
Retrospect of previous contents, especially from the intro
%NTS:::EXTIRPATED means locally extinct

This thesis treated extinction from a one dimensional logistic system in chapter 1, extinction from a coupled logistic system in chapter 2, and invasion into a coupled logistic system in chapter 3. The coupled logistic system of chapters 2 and 3 acts like a neutral model in the Moran limit, and the transition to this limit was characterized. Chapter 3 also looks at dynamics and steady state distributions of a single species in a neutral model. %NTS:::chapter numbers
More detailed discussions and conclusions specifically related to these topics can be found therein. 

Here I would like to address the general questions brought forward in the introduction. 
These are: how long will a species exist on its own, among others, and what this implies for biodiversity. 
Furthering the discussion of biodiversity, I want to consider immigration probabilities and repeated invasions (into the Moran model). 

exponential implies surviving is fine, but even a 1d stable system mkinght not ob=serve that exactly
what approximations workk (and when)
comment on niche vs neutral - neutral?, for parsimony - I can't really conclude
maintenance of biodivefrsity

\iffalse
The major questions of this thesis are: What are the probability and timescale of a single species extinction in an ecosystem? How should the probability and MTE be calculated? Inspired by problems of biodiversity, what is the mean time to fixation of two competing species? Conversely, what is the probability and timescale of invasion of a second species into an ecosystem occupied by a first? 
The structure of the thesis is as follows. 
First, I use the exact techniques mentioned above and introduced more completely in sections 2.3 and 2.4 to investigate a one dimensional logistic system, comparing the influence of the linear and quadratic terms to the quasi-steady state distribution and the MTE. %NTS:::chapter/section number
I find that those species with high birth and death rates, and those for whom competition acts to increase death rate rather than reduce their birth rate, tend to go extinct more rapidly. %CONCLUSION
With the simplicity of this test system I explore the applicability of various common approximation techniques. 
I conclude the Fokker-Planck approximation works well close to the deterministic fixed point, but incorrectly estimates the scaling of the extinction time with system size. The WKB approximation performs better, but misidentifies the prefactor to the exponential scaling. %CONCLUSION
The exact techniques and the approximations together make up chapter 1, regarding a one dimensional system. %NTS:::chapter number
This chapter is being prepared as a paper for publication \cite{Badali2018a}. 
The natural extension from a one dimensional logistic is to couple two such systems together; this arrives at the two dimensional generalized Lotka-Volterra system and is the subject of the next chapter, chapter 2. %NTS:::chapter number
%First a symmetric system is investigated, and t
The mean time to fixation is used as a tool to diagnose the longevity of the two interacting species. 
The overlap of their ecological niches is the parameter that controls the transition between effective coexistence and rapid fixation. 
I determine that two species will effectively coexist unless they have complete niche overlap, even if they have only a slight niche mismatch. %CONCLUSION
%Next the corresponding asymmetric model is explored. 
Along with the MTE, my analysis uncovers a typical route to fixation, or rather a lack of a typical route, the discussion of which wraps up this chapter. %kinda CONCLUSION
The final chapter introducing novel research, chapter 3, extends the scope of this thesis to invasion of a new species into an already occupied niche. %NTS:::chapter number
I calculate the probability of a successful invasion as a function of system size and niche overlap. 
Then the MTE conditioned on the success of the invasion is analyzed. 
I discover that the closer the invader is to having complete niche overlap with the established species, the less likely it is to successfully invade, and the longer an invasion attempt will take before it is resolved. %CONCLUSION
Once these timescales are developed, I regard the Moran model modified to account for repeated invasions of the same species. 
%This is compared with some steady state numerical results from Kimura. 
%I demonstrate that, with system size $K$ and relevant immigrant probability $g$, an immigration rate of $1/K g$ is the critical value for determining the qualitative abundance distribution. %CONCLUSION
I identify the critical value of the immigration rate above which a species will have a moderate population size and below which the population is either large or largely absent in its contribution to the abundance distribution. %CONCLUSION
Chapter 2 and half of chapter 3 together form another paper being reviewed for publication \cite{Badali2018}. %NTS:::chapter numbers
The conclusions chapter covers a variety of topics: I explore applications and extensions of the results arrived at in this thesis; I address the central problems introduced in this preliminary chapter and draw some conclusions informed by my results; and I suggest next steps for this research, both continuations and implementations to novel situations. 
%NTS:::somewhere need to put in who contributed to what.
%some good verbs: confirm find infer establish identify discover demonstrate show
\fi
\iffalse
Big Questions:
How long will a single species exist with only intraspecies interactions?
What mathematical techniques are effective to model such an extinction? 
How long will a species exist with intra and interspecies interactions? That is, how long will two species coexist given some niche overlap? In particular, how does it transition from the effective coexistence of exponential scaling of MTE with carrying capacity to the relatively fast extinction of algebraic scaling as found in the Moran model/limit? 
For an ecosystem with an already established species, what is the probability of success and the timescale of an invasion attempt? How do these probabilities and times depend on niche overlap? 
With repeated immigration of a species, how will that species be distributed in a [neutral] system? In particular, how does the distribution depend on the immigration rate? 
What is the timescale of species transient existence in a neutral model with repeated immigrants? 
Is v=1/gN or gv=1/N the same as a model with v’=gv and g=1? - looks like it should be
\fi



\section{Next steps for the research}
%NTS:::in INTRO chapter, mention that my interest is in the hard problems far from equilibrium; not just stochastics (which are already more complicated than deterministics) but the rare events like first passages
%NTS:::in INTRO, "minimal working model" rather than null model
%NTS:::in CH1, point out that inverting the matrix gives perfect match with true results
%NTS:::in CH1, Langevin is the same as Fokker-Planck (though it is often done even worse)
%NTS:::comment somewhere (intro chapter) about what defines fitness???
I will now indulge in some speculation. 
In this section I present some straightforward extensions of my research. 
Also included are some more extensive next steps. 

%From the first chapter: approximations
Not much more could be done with the approximations, unless I omitted a major one. 
Biophysics ias a field is dnamic and as such I would not be surprised if a new technique gains populatirty in the next few years. 
Likely the technique already exists and has not made its way to our discipline. 
It could be in far from equilibrium condensed matter or high energy physics, it might be evolutionary game theory, or already commonly employed in linguistics, economics, graph theory, or the more mathematical side of stochastic processes. 
Martingales are soluble, so perhaps the next approach will be to map everything to a martingale or a convolution of martingales. 
Artificial intelligence is also trendy at the time of writing, and the many layered neural nets that are becoming available could easily be turned to stochastic processes, with each neuron representing one population state of the system. 
Current routinely used nets have ~XXXX neurons, which corresponds to a sqrt(XXXX) carrying capacity. 

%From the first chapter: biology
%In addition, t
The content of the first chapter has a couple extensions I would like to try. 
The ``hidden parameters'' were those that did not show up in the deterministic analogue of the system which nevertheless affect the stochastic dynamics. 
Suppose we write the system as
\begin{align*}
	b(n) &= r\,n + f(n) \\
	d(n) &= r\,n^2/K + f(n)
\end{align*}
for arbitrary $f(n)$. I should like to see how $f$ affects the stochastic measurables like MTE. 
Writing $f$ as a polynomial, I predict the higher orders will have a lesser and lesser effect, and that there will be some conditions placed on $f^{(k)}(0)/k!$, the $k$th coeffient. 
Furthermore, I investigated the deterministic equation $\dot{x} = r\,x(1-x/K)$. I suspect any concave down system will behave similarly \cite{Strogatz?}. 
It is unclear what the effects of other forms would be. 
For instance, Kamenev and others studied inclusion of the Allee effect \cite{Kamenev?}, which to my mind would give extinction time scaline with an effective carrying capacity equal to the difference between the stable and unstable fixed points. 

%chapter 2 stuff - see if the Moran condition is ruined by adding anything (already measure zero); include evolution, possibly which would stabilize, possibly leading to greater likelihood of Moran
In chapter 2 I regarded how the scaling of the extinction time changes as niche overlap increases from independence to $a=1$, at which point the MTE is that of the Moran model. 
It is only in this one limit that coexistence is not long-lived (with the large population caveat examined in section 2.7); with the size of phasespace, this one point seems unlikely, being of measure zero. 
A next step would be to account for mutation and evolution in this model. 
If a mutant strain arises from a single population, it is likely to have a very similar niche to the wildtype, hence a large niche overlap. 
And from there, what would the forces of evolution dictate? I would continue to constrain the system to the case wherein neither species has an intrinsic fitness advantage. 
The niche overlap could be allowed to evolve, but it would be more sensible to instead define a niche space (perhaps 1D, perhaps multidimensional) in which the niche of each species is defined (likely with a Gaussian distribution \cite{MacArthur}). 
The desire to increase fitness might have a stabilizing on the system, encouraging differing niches. Or it might keep the niche overlap close to unity, which would go far to explaining why the Hubbell model has had such success. 
In any case, it would be insightful to see whether fitness considerations act to make the already-unlikely Moran limit entirely untenable, or whether this limit is actually a natural consequence of evolution. 

%chapter 2 stuff - selection (weak and strong)
%NTS:::in my breaking the symmetries section I didn't consider breaking the "r"s. I should (or else do it now (or else talk about it now))
A natural extension of my work would be to include selection, explicit fitness advantages, to the system. 
I have not done so already because selection has been treated many times in many ways \cite{a bunch???} and actually tends to act to simplify the system. 
The higher fitness species rapidly fixates. 
Nevertheless there are some advantages to how I treat stochastic systems that would aid the analysis of systems with selection. 
In the independent limit any fitness advantage should be irrelevant, so there will at least be a transition between coexistence in the independent limit and rapid fixation otherwise. 
What is more, most models with selection, including those of Kimura \cite{Kimura??} and Moran \cite{Moran with selection??} [and others???] must make the assumption that the effect of selection is weak, typically much less than $1/K$. 
Inverting the transition matrix, as I do in chapter 2, is arbitrarily precise and so does not suffer from any requirement of such an assumption. It can treat the range of selection, from weak to strong. 
Not only does this give access to regimes not normally considered, it provides a way to verify the small selection results that rely on approximation. 
%one more sentence?!?
In the literature there is a debate about selective sweeps \cite{Jensen2014} that requires more careful treatment, with fewer assumptions; the method employed in chapter 2 of this thesis would be ideal to explore the differences between hard and soft selective sweeps. 
Selection can also be incorporated into the invasion dynamics of chapter 3. 

%chapter 2-3 stuff - environmental noise
One extension applicable to all my work, that I did not account for, is the inclusion of environmental noise. 
This has a fundamentally different action on the system. 
Whereas demographic stochasticity comes from the discrete nature of population sizes and therefore can be mapped onto a transition matrix or graph (as in figure 1.1),%NTS:::figure number
environmental noise is continuous, especially for the phenomenological parameters I use, which cannot be fixed to any one cause, let alone a discrete one. 
Such an advancement in my research would involve the abandonment of the transition matrix, and likely would require the use of Fokker-Planck, which I am dubious about, based on my findings in chapter 1. 

%chapter 3 stuff - check what happens to Hubbell/abundance with incomplete niche overlap
%In chapter 3 I focused on a single species immigrating into a small system. 
In chapter 3 I considered the probability and timescales of a single individual attempting to invade a system with an established species, and observed the effect of niche overlap. 
I then looked at the steady state population distribution and conditional exit times of the focal species in the case of complete niche overlap, via the Moran model. 
This Moran model was equivalent to the system employed by Hubbell \cite{Hubbell2001}, albeit with different questions being asked. 
However, I could have also asked the questions of Hubbell in the case of incomplete niche overlap. 
Namely, if species do not occupy the same niche, what should we expect the species abundance curves to look like? 
With lessened pressure of competing for exactly the same niche I showed that an invader can more easily invade a system, and in lesser time, regardless of its ultimate success or failure. 
This suggests that a Hubbell model with only partial niche overlap would certainly have more species than the classic Hubbell model with $a=1$. 
It is unclear how the abundance curves would change. 
The Hubbell model has its total population size $K$, but if species occupy different niches they could in principle have different carrying capacities. 
Therefore the abundance curve would be influenced by the distribution $K$ is drawn from. 
In effect this change of niche overlap being partial and carrying capacities drawn from a distribution would move the model of Hubbell closer to those models of niche apportionment \cite{??niche apportionment??}. 
Insofar as no species would have an explicit fitness advantage this new model would be neutral, but with its differing $K$s it would also be a niche model, hence a hybrid of the two. 
There is a body of work for hybrid models \cite{??hybrid niche neutral??} against which one could compare my proposed incomplete niche overlap Hubbell model. 

%chapter 3 stuff - coalescent with incomplete niche overlap
Just as the Hubbell model is used to make sense of current extant species, coalescent theory uses a Moran or Kimura model to understand paleontological data and other data of the past \cite{??coalescent??}.  
It seeks to find the time to coalescence, which means the time (in the past) when a common ancestor existed for two variants of a species (in the present). 
The classic model assumes the variants behave the same - in my language, they occupy the same niche, and have niche overlap of one. 
The theory has been modified to include explicit fitness advantage \cite{???} but the unbiased case of incomplete niche overlap has not been tried. %as far as I know/to the best of my knowledge. 
Seeing as decreasing the niche overlap from unity acts to increase the timescales of the system, as I have shown in this thesis, using my results to extend coalescent theory when appropriate would similarly suggest that variants diverged longer ago than the classic theory predicts. 

The techniques applied in this thesis, specifically the use of a truncated transition matrix being inverted to solve for the first passage times exactly, can easily be applied to other low species number systems. 
The calculation for each point in parameter space is not lengthy; the main constraint is RAM, rather than time, and so a larger memory computer could deal with problems with larger carrying capacities or more species. %NTS::: can expand these past two sentences into a whole paragraph
For example, in chapter 2 I considered a two dimensional generalized Lotka-Volterra system to explore competition between two species. 
The Lotka-Volterra system has been generalized to any dimension, any number of species. 
A judicious choice of parameters will recreate the predator-prey dynamics from which the generalized Lotka-Volterra system gets its name. 
Assaf \cite{} analyzed the predator-prey system using a clever rotating reference frame and the WKB approximation. 
It would be nice to get the arbitrarily correct data to compare to the approximate results of their research. 
Furthermore, the deterministic system is marginally stable, with trajectories orbiting a zero-eigenvalue fixed point. As such, I expect an algebraic time to extinction, as well as a stronger dependence on initial conditions. 
However, the period depends on the value of the Lyapunov energy associated with the orbit. Thus as it ``diffuses'' tangentially to the orbit its characteristic time scale will change, which may complicate the analysis. 
Nevertheless, the data that can be generated by inverting a truncated transition matrix to solve the extinction time to arbitrary accuracy would be correct, regardless of the analysis. 

From the two dimensional Lotka-Volterra system one arrives at the predator-prey system by choosing the parameters correctly. 
One can also move to the third dimension, in order to account for a third species in the system. 
This allows for the investigation of many systems of interest, with much more diversity. 
The simplest extension in this regard would be to have three species all with overlapping niches. 
I could observe how a species whose niche is situated between those of two others (such that they each overlap with the first species but not with each other) would go extinct more readily as the overlap of the encroaching species is increased. 

A famous fun model of three interacting species has ones that %RPS - not just fun, also with pairwise competition of bacteria a la Gore

%NTS::: SIR, 3 species fewer resources highlighted by McGehee, chaos

%NTS:::suggestions in this paragraph
%applying the results to things %As with all my research, there is the obvious extension of applying my results to specific real biological systems, finding a way to estimate the phenomenological parameters based on measurable evidence, and then making preditions. 
%3D+ (predator-prey, SIR, etc., chaos?) - see above; niches, $K$ drawn from distribution
%could couple eco and evo, to move the niches (or their overlap) based on competition with the other species in the system or with some ``hidden'' resource parameters
%other possibilities?





